Energy and exergy analysis of a cruise ship

### 2018-06-05 Manuscript/Manuscript.tex
Author update
Updates after meeting - Part 2

\author[EPFL]{Francesco Baldi}
\author[Linnaeus]{Fredrik Ahlgren}
\author[DTU]{Tuong-Van Nguyen}
\author[CTU]{Cecilia Gabrielii}
\author[CTU]{Karin Andersson}

\address[EPFL]{Industrial Process Energy Systems Engineering (IPESE), \'{E}cole Polytechnique F\'{e}d\'{e}rale de Lausanne, 1950, Sion, Switzerland}
\address[Linnaeus]{Kalmar Maritime Academy, Linnaeus University, Kalmar, Sweden}
\address[DTU]{Department of Mechanical Engineering, Technical University of Denmark, Lyngby, Denmark}
\address[CTU]{Department of Mechanics and Maritime Sciences, Chalmers University of technology, Gothenburg, Sweden}

@@

The cruise industry have also received much attention for its environmental impact other than in terms of GHG emissions. Cruise ships were shown to have a remarkable negative impact on both air \cite{Poplawski2011} and water \cite{Butt2007} quality, and were also pointed as sources of environmental damage from a wider perspective \cite{Brida2009}. The fact that cruise ships tend to operate extensively in highly populated and environmental sensitive areas makes the situation all the more worth of attention.

Altogether, these conditions present a challenge to the shipping companies who attempt to reduce their fuel consumption, environmental impact, and operative costs. A wide range of fuel saving solutions for shipping are available and partially implemented in the existing fleet, both from the design and operational perspective; several specific studies have been conducted on these technologies, and a more detailed treatise would be out of the scope of this work. In this context, it has been acknowledged that the world fleet is heterogeneous, and measures need to be evaluated on a ship-to-ship basis \cite{Bouman2017}. In this process, a deeper understanding of energy use on board of the specific ship is vital.

@@

Most authors focused on the propulsion part of the problem, as this is often the most relevant energy demand on board. Shi et al. proposed a modeling approach for predicting ship fuel consumption of a cargo/passenger ferry \cite{Shi2009}; Theotokoats and Tzelepis applied a similar procedure to the case of a Handymax product carrier \cite{Theotokatos2015}, similarly to Tillig et al., who also added the dynamic element to their predictive model \cite{Tillig2016}.

@@

More generally, several authors have highlighted the importance of a detailed knowledge of the ship's operational profile in order to appropriately assess and optimize possible alternatives for improving ship energy efficiency \cite{Banks2013}. Coraddu et al., started from a statistical analysis of measured ship operations and used the aggregated data, together with a computational model of the vessel, to provide a better prediction of the actual ship's operational efficiency \cite{Coraddu2014}. The importance of considering the operational profile was also showed in the case of the optimization of engine-propeller interaction \cite{Baldi2015c}, in the process of retrofitting existing systems \cite{Baldi2015b,Choi2013}, and in ship design \cite{Ghassemi2017,Solem2015}.

% Heat availability
Heat demand is rarely a subject of concern on board ships, with some notable exceptions. This is due to a combination of generally low demand and high availability from the waste heat of the engines. In a previous publications by the authors, for instance, it is shown that in the case of a product tanker, although the heat demand was estimated to account for roughly 20\% of the total energy demand of the ship on a yearly basis, it only contributes to 4.1\% of the total fuel consumption (contribution of the auxiliary boilers), while the rest of the demand is fulfilled using waste heat \cite{Baldi2015a}.

@@

The potential uses for waste heat on board are not limited to improving the efficiency of the power plant. Waste heat is commonly used for fulfilling on board heat demand for spatial heating and freshwater generation \cite{Molland2011,Baldi2015a,Mccarthy1990};  Balaji et al. proposed the use of waste heat for ballast-water systems \cite{Balaji2012}; Salmi et al. suggested its use for adsorption refrigeration systems \cite{Salmi2017}. A detailed review of potential uses for waste heat from marine engines is presented by Shu et al \cite{Shu2013}.

@@

% Exergy analysis
Exergy analysis provides a more accurate estimation of the potential for energy recovery on board. The application of exergy analysis to the case of ship energy systems is, however, still limited. Dimopoulos et al. showed how the process optimizing the WHR system of a container ship can be more efficient if exergy efficiency, rather than energy efficiency, is used as the target of the optimization \cite{Dimopoulos2012}; Baldi et al. also analyzed the exergy flows on board of a product tanker, showing that this allows having a more accurate understanding of what parts of the system show potential for improvement \cite{Baldi2015a}; similar results were obtained by Marty et al., who focused on the power plant of a cruise ship \cite{Marty2016}. Koroglu et al. made a step futher also including advanced exergy analysis in their study \cite{Koroglu2017}.

@@

Given the existing limitations of the current available information in scientific literature highlighted in the previous section, the objective of this work is the following:
\begin{itemize}
	\item Analyse the demand of a cruise ship in terms of propulsion power, electric power and heat, based on operational measurements. The lack of existing information with reference to the heat demand is considered a particular element of novelty when compared to existing literature in the subject.
	\item Analyse the current efficiency of the system, and potential ways to improve it, by means of applying exergy analysis.
	\item Provide reference operational conditions for further use in the field of energy systems optimization.
\end{itemize}

@@

In this paper, we present the application of energy and exergy analysis (described in section \ref{sec:met:energyExergy}) to a cruise ship. This is done for one specific case study vessel, that is described in detail in Section \ref{sec:met:case}; in Section \ref{sec:met:gathering} we present the specifics of the available information, in particular the measured data and the technical documentation; the specific assumptions and methods used for processing the available information into  energy and exergy flows are specified in Section \ref{sec:met:processing}. In these regards, the estimation of the heat demand is treated in a separate section (Sec. \ref{sec:met:heat}) as it constitutes a particularly challenging task and it is one of the central aspects of this article.

@@

\emph{Energy} can be stored, transformed from one form to another (e.g. heat to power) and transferred between systems, but can neither be created nor destroyed (conservation law)~\textbf{Bejan et al.}. The system under study is the \emph{ship energy system} and is thus taken as control volume. The energy balance can be expressed as:

@@

The energy flow associated with a material stream is calculated as the sum of the physical and chemical enthalpies, and kinetic and potential energies are neglected.    The physical energy is taken as the relative enthalpy, as underlined in~\textbf{Kotas et al.}, and the chemical energy is taken as the lower/higher heating value. The environmental conditions taken for the present analysis are the ambient pressure and seawater temperature.

@@

The left-hand side term represents, on a time rate basis, the exergy associated with the fuel and air. The right-hand side term denotes the exergy of the waste streams, the exergy transfers with heat and power, and the exergy destroyed in the ship system. Kinetic and potential exergies are neglected, and the exergy of a given material stream is derived as the sum of the physical and chemical exergies. The chemical exergy is calculated based on the reference environment of \textbf{Szargut (1989)}, and its value is approximatively equal to the higher heating value for hydrocarbon fuels. The exergy destruction can be calculated from the Gouy-Stodola theorem~\textbf:{Kotas (1995)}.

@@

The following indicators are used to evaluate the system performance:
\begin{itemize}
	\item the exergy efficiency $\varepsilon$, defined as the ratio between the exergy product and fuel of a given component or system
	\begin{equation} \varepsilon = \frac{\dot{E}_{\mathrm{p}}}{\dot{E}_{\mathrm{f}}} \end{equation}
	\item the efficiency defect $\lambda$, presented in the work of Kotas, defined as the fraction of the total exergy input destroyed in the successive irreversible processes
	\begin{equation} \lambda = \frac{\dot{E}_{\mathrm{d}}}{\dot{E}_{\mathrm{in,tot}}} \end{equation}
	\item  the irreversibility share $\delta$, suggested in the work of Tsatsaronis, defined as the ratio between the exergy destroyed in the $i$-th component in relation to the exergy destroyed in the entire system
	\begin{equation} \delta_i = \frac{\dot{E}_{\mathrm{d},i}}{\dot{E}_{\mathrm{d,tot}}} \end{equation}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.95\linewidth]{Figures/ShipLayout}
	\caption{Layout of the case study energy systems}
	\label{fig:shipSystems}
\end{figure}

@@

@@

@@

The setup of the engines and their connections to the other ship systems are summarized in Figure \ref{fig:DieselEngines} for the main and auxiliary engines, respectively. The most relevant operative values are provided in Tables \ref{tab:ME_values} and \ref{tab:AE_values} and the main energy flows are listed in Tables \ref{tab:ME_flows} and \ref{tab:AE_flows}. It should be noted that, in the case of the main engines, the engine power output was not measured and needed to be estimated. In this work, we calculated the engine power output based on measurements of engine fuel rack position (used as a proxy of the mass of fuel injected per cycle) and of the engine speed:
\begin{eqnarray}
\dot{m}_{fuel} & = & \dot{m}_{fuel,des} \left(a_0 + a_1 \frac{frp}{frp_{des}}\right) \frac{\omega}{\omega_{des}} \\
\eta_{ME} & = & a_0 + a_1 \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} + a_2 \left( \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} \right)^2 \\
\dot{W}_{ME} & = & \dot{m}_{fuel} \eta_{ME} LHV
\end{eqnarray}

@@

In the case of the auxiliary engines, instead, direct measurements of the power generated by each engine were available.

For both auxiliary and main engines, part of the relevant internal variables are not measured (particularly with relation to the bypass valves) and this required to determine them based on modeling the heat and mass balance of the engine. 
\begin{align}
	\dot{m}_{air,comp} & =  \dot{m}_{air,cyl} + \dot{m}_{air,bp} \label{eq:mb:comp} \\ 
	\dot{m}_{eg,turb} & = \dot{m}_{eg,cyl} + \dot{m}_{air,bp} \label{eq:mb:turb} \\
	\dot{m}_{eg,cyl} & = \dot{m}_{air,cyl} + \dot{m}_{fuel,cyl} \label{eq:mb:cyl} \\
	\dot{m}_{air,comp} \Delta h_{comp} & = \dot{m}_{eg,turb} c_{p,eg} (T_{turb,in} - T_{turb,out}) \eta_{mech} \label{eq:eb:turbocharger} \\
	\dot{m}_{eg,turb} c_{p,eg} (T_{turb,in} - T_0) & = \dot{m}_{air,bp} c_{p,air} (T_{comp,out} - T_0) + \dot{m}_{eg,cyl} c_{p,eg}  (T_{cyl,out} - T_0) \label{eq:eb:bpmerge}
\end{align}
Where equations \ref{eq:mb:comp} to \ref{eq:mb:cyl} represent the mass balances of the bypass split, bypass merge, and cylinder respectively, while equations \ref{eq:eb:turbocharger} and \ref{eq:eb:bpmerge} represent the energy balances of the turbocharger and of the bypass merge, respectively.

Values for temperatures, mass flows and energy flows for different engine loads are provided in the \ref{sec:appendix:engines}, for both main and auxiliary engines.

\subsubsection{Cooling systems}

The ship's cooling systems are designed similarly to most ships and divided in two cooling systems of two separate temperature levels: the high temperature (HT) cooling systems (70-90$^o$C) and the low temperature (LT) cooling systems (40-60$^o$C). The cooling systems provide the necessary cooling to the engines, to the steam systems and to all other components on board. A representation of the ship cooling systems is provided in Figure \ref{fig:generalsystemssimplified}.

All values not directly measured in the cooling systems were calculated based on mass and energy balances. Mass flows were determined based on measurements of the pressure in the system and on the performance curves of the cooling pumps as provided in the engine project guides.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/GeneralSystemsSimplified}
	\caption{Layout of the ship heating and cooling system}
	\label{fig:generalsystemssimplified}
\end{figure}

@@

The total electric power demand of the system is easily estimated as the sum of the power generated by the four auxiliary engines, that is directly measured on the electrical side of the generators and has a rather high accuracy
\begin{equation}
P_{el,tot} = \sum_{1}^{4} P_{AE_i}
\end{equation}

The ship is equipped with a number of different systems, including lighting, navigational systems, pumps and compressors, etc. However, none of the individual contributions is directly measured, hence estimations of the energy demand of individual consumers need to be based on indirect measurements. In this paper we identified methods for the estimation of the contribution of two consumers: the HVAC systems and the bow thrusters.

\subsection{Heat demand} \label{sec:met:heat}

@@

The main view of the ship heating and cooling systems is presented in Figure \ref{fig:ship:general}.

@@

@@

Improving the performance of a ship requires a detailed analysis of the energy demands to optimize the dimensioning and operation of the technologies and processes to install on-board. It requires the analysis of the hourly, daily and seasonal variations of the heat/power consumption, operating (e.g. ship speed) and external conditions (e.g. air temperature). However, gathering and processing those data may be time-consuming because of the high number of points to assess. It is therefore desirable to reduce this large quantity of information while keeping the relevant details on the relations between the variables of interest. 
One way is to represent these yearly profiles in a limited number of representative (typical) periods, avoiding the repetition of similar data sets. Clustering consists of grouping such sets in a single group (cluster) so that the items in the same group are more like each other than to those belonging to other groups. Several algorithms are introduced in the literature to partition these datasets. The one selected in this work is the Lloyd’s (k-means) algorithm, which is a non-deterministic method computationally faster than conventional hierarchical tools. The approach applied by Fazlollahi et al. was used to select the optimum number of clusters (Nk). It (Nk) should be as low as possible for data handling purposes, while preserving a high accuracy of the retrieved data. It builds on the calculation of three criteria: 
\begin{enumerate}
	\item the average intra-cluster distance to assess the density of each cluster, preferably small;
	\item the average inter-cluster distance to assess the distance between each, preferably high; 
	\item the expected square error, which is a statistical measure suggested by Pham et al.. 
\end{enumerate}

In parallel, the cluster quality was systematically evaluated by calculating the following performance indicators:
\begin{enumerate}
	\item the profile deviation (deviation between original and typical period profiles),$\sigma_{profile}$;
	\item the deviation from the load duration curve of the average values of each period, $\sigma{CDC}$;
	\item the relative error in load duration curve deviation $ELDC$;
	\item the maximum duration load curve difference $\Delta_{MLDC}$;
	\item the number of periods with relative errors $\Delta_{prod}$ higher than 7\%.
\end{enumerate}

@@

Figure \ref{fig:vship_hist} represents the distribution of the ship speed. The ship operates for almost  the entire year according to a fix schedule, while there are a couple of periods (particularly during the summer months) when the ship operates at higher speed. In Figure \ref{fig:opMode_pie} it can be also observed that the ship spends a relatively large amount of time in port. These remarks suggest that, in spite of the high installed power of the main engines, the energy demand for propulsion is expected to be relatively low.

@@

The ship energy demand is first subdivided among  The evolution of the demands of propulsion, electric power, and heat over time during a typical voyage is shown in Figures \ref{fig:Demand_vs_time_W} and \ref{fig:Demand_vs_time_S}, representing a winter and a summer day respectively. As expected, the heating demand in winter is higher than in summer [2800-4500 kW vs 1100-2900 kW], as a consequence of the reduced need for compartment heating. On the other hand, the electric power demand behaves inversely, ranging around 1900 kW during the reference winter day (peaks are connected to the use of thrusters in port for maneuvering) and around 2300 kW during the reference summer day, where the difference is mostly associated to the demand of the HVAC compressors. Looking at the yearly cumulated demand, nearly 50\% of the total energy demand is related to ship propulsion (see Figure \ref{fig:Pie_EnergyDemand}), while the remaining portion is approximately equally split between electric energy and heat demand.

@@

Heat recovery has a large impact on the overall heat demand (see Fig.  \ref{fig:Pie_HeatGeneration}). The exhaust gas boilers and the heat recovery on the HT-water fulfill almost 75\% of the yearly demand for heating, leaving only the remaining 25\% to be provided by oil-fired auxiliary boilers. The contribution of the HRSG and the HTHR is substantially constant throughout the year (Fig. \ref{fig:TimeSeries_HeatGeneration}), while the auxiliary boilers are used to cover the peaks and the winter increased baseload.

@@

Given the fact that many engines are operated at low load (both main and auxiliary engines), the system would benefit from both an electrification of the system and from the installation of batteries. The system electrification was explored by the authors \cite{Baldi2016b} and showed a relevant potential from both an environmental and an economic point of view. The installation of batteries was also considered in a previous publication by the authors \cite{Baldi2017} and proved the potential for yearly savings of 1-2\%. In this second case, however, it should be noted that the economic performance was not ensured.

The availability of waste heat that is not already recovered by the system suggests that the system could benefit from the installation of a Heat-to-power WHR. This possibility was tested by the authors both in the case that no additional retrofit is performed \cite{Ahlgren2016,Mondejar2017}, with estimated savings of 22\% of the auxiliary power demand for a reference voyage, and in the case of a full integration with the rest of the system \cite{Baldi2017}, with estimated savings of up to 6\% of the total fuel consumption for a year of ship operations.

@@

\section{Engine modelling, figures and tables} \label{sec:appendix:engines}

@@

@@

>>>>

\author[EPFL]{Francesco Baldi}
\author[Linnaeus]{Fredrik Ahlgren}
\author[DTU]{Tuong-Van Nguyen}
\author[LTU]{Marcus Thern}
\author[CTU]{Karin Andersson}

\address[EPFL]{Industrial Process Energy Systems Engineering (IPESE), \'{E}cole Polytechnique F\'{e}d\'{e}rale de Lausanne, 1950, Sion, Switzerland}
\address[Linnaeus]{Kalmar Maritime Academy, Linnaeus University, Kalmar, Sweden}
\address[DTU]{Department of Mechanical Engineering, Technical University of Denmark, Lyngby, Denmark}
\address[LTU]{Energy Sciences, Lund University, Lund SE-22100, Sweden}
\address[CTU]{Department of Mechanics and Maritime Sciences, Chalmers University of technology, Gothenburg, Sweden}

@@

The cruise industry has also received much attention for its environmental impact other than in terms of GHG emissions. Cruise ships were shown to have a remarkable negative impact on both air \cite{Poplawski2011} and water \cite{Butt2007} quality, and were also pointed as sources of environmental damage from a wider perspective \cite{Brida2009}. The fact that cruise ships tend to operate extensively in highly populated and environmental sensitive areas makes the situation all the more worth of attention.

Altogether, these conditions present a challenge to the shipping companies that attempt to reduce their fuel consumption, environmental impact, and operative costs. A wide range of fuel saving solutions for shipping are available and partially implemented in the existing fleet, both from the design and operational perspectives; several specific studies have been conducted on these technologies, and a more detailed treatise would be out of the scope of this work. In this context, it has been acknowledged that the world fleet is heterogeneous, and measures need to be evaluated on a ship-to-ship basis \cite{Bouman2017}. In this process, a deeper understanding of energy use on board of the specific ship is vital.

@@

Most authors focused on the propulsion part of the problem, as this is often the most relevant energy demand on board. Shi et al. \cite{Shi2009} proposed a modeling approach for predicting ship fuel consumption of a cargo/passenger ferry; Theotokoats and Tzelepis \cite{Theotokatos2015} applied a similar procedure to the case of a Handymax product carrier, similarly to Tillig et al. \cite{Tillig2016}, who also added the dynamic element to their predictive model.

@@

More generally, several authors have highlighted the importance of a detailed knowledge of the ship's operational profile in order to appropriately assess and optimize possible alternatives for improving ship energy efficiency \cite{Banks2013}. Coraddu et al. \cite{Coraddu2014} started from a statistical analysis of measured ship operations and used the aggregated data, together with a computational model of the vessel, to provide a better prediction of the actual ship's operational efficiency . The importance of considering the operational profile was also showed in the case of the optimization of engine-propeller interaction \cite{Baldi2015c}, in the process of retrofitting existing systems \cite{Baldi2015b,Choi2013}, and in ship design \cite{Ghassemi2017,Solem2015}.

% Heat availability
Heat demand is rarely a subject of concern on board ships, with some notable exceptions. This is due to a combination of generally low demand and high availability from the waste heat of the engines. In a previous publications by the authors \cite{Baldi2015a}, for instance, it is shown that in the case of a product tanker, although the heat demand was estimated to account for roughly 20\% of the total energy demand of the ship on a yearly basis, it only contributes to 4.1\% of the total fuel consumption (contribution of the auxiliary boilers), while the rest of the demand is fulfilled using waste heat.

@@

The potential uses for waste heat on board are not limited to improving the efficiency of the power plant. Waste heat is commonly used for fulfilling on board heat demand for spatial heating and freshwater generation \cite{Baldi2015a,Molland2011,Mccarthy1990};  Balaji et al. \cite{Balaji2012} proposed the use of waste heat for ballast-water systems ; Salmi et al. \cite{Salmi2017} suggested its use for adsorption refrigeration systems. A detailed review of potential uses for waste heat from marine engines is presented by Shu et al \cite{Shu2013}.

@@

% Exergy analysis
Exergy analysis provides a more accurate estimation of the potential for energy recovery on board. The application of exergy analysis to the case of ship energy systems is, however, still limited. Dimopoulos et al. \cite{Dimopoulos2012} showed how the process optimizing the WHR system of a container ship can be more efficient if exergy efficiency, rather than energy efficiency, is used as the target of the optimization; Baldi et al. \cite{Baldi2015a} also analyzed the exergy flows on board of a product tanker, showing that this allows having a more accurate understanding of what parts of the system show potential for improvement ; similar results were obtained by Marty et al. \cite{Marty2016}, who focused on the power plant of a cruise ship. Koroglu et al. \cite{Koroglu2017} made a step futher also including advanced exergy analysis in their study.

@@

Given the existing limitations of the current available information in scientific literature highlighted in the previous section, the objective of this work is the following:
\begin{itemize}
	\item Analyse the demand of a cruise ship in terms of propulsion power, electric power and heat, based on operational measurements. The lack of existing information with reference to the heat demand is considered a particular element of novelty when compared to existing literature in the subject.
	\item Analyse the current efficiency of the system, and potential ways to improve it, by means of applying exergy analysis.
	\item Provide reference operational conditions and typical operational days for further use in the field of energy systems optimization.
\end{itemize}

@@

In this paper, we present the application of energy and exergy analysis (described in Section \ref{sec:met:energyExergy}) to a cruise ship. This is done for one specific case study vessel, that is described in detail in Section \ref{sec:met:case}; in Section \ref{sec:met:gathering} we present the specifics of the available information, in particular the measured data and the technical documentation; the specific assumptions and methods used for processing the available information into  energy and exergy flows are specified in Section \ref{sec:met:processing}. In these regards, the estimation of the heat demand is treated in a separate section (Section \ref{sec:met:heat}) as it constitutes a particularly challenging task and is one of the central aspects of this article.

@@

\emph{Energy} can be stored, transformed from one form to another (e.g. heat to power) and transferred between systems, but can neither be created nor destroyed (conservation law) \cite{Bejan1996}. The system under study is the \emph{ship energy system} and is thus taken as control volume. The energy balance can be expressed as:

@@

The energy flow associated with a material stream is calculated as the sum of the physical and chemical enthalpies, and kinetic and potential energies are neglected. The physical energy is taken as the relative enthalpy, as underlined in \cite{Kotas2013}, and the chemical energy is taken as the lower/higher heating value. The environmental conditions taken for the present analysis are the ambient pressure (1.01 bar) and seawater temperature (measured) .

@@

The left-hand side term represents, on a time rate basis, the exergy associated with the fuel and air. The right-hand side term denotes the exergy of the waste streams, the exergy transfers with heat and power, and the exergy destroyed in the ship system. Kinetic and potential exergies are neglected, and the exergy of a given material stream is derived as the sum of the physical and chemical exergies. The chemical exergy is calculated based on the reference environment of \cite{Szargut1989}, and its value is approximatively equal to the higher heating value for hydrocarbon fuels. The exergy destruction can be calculated from the Gouy-Stodola theorem \cite{Kotas2013}.

@@

The following indicators are used to evaluate the system performance:
\begin{itemize}
	\item the exergy efficiency $\varepsilon$, defined as the ratio between the exergy product and fuel of a given component or system
	\begin{equation} \varepsilon = \frac{\dot{E}_{\mathrm{p}}}{\dot{E}_{\mathrm{f}}} \end{equation}
	\item the efficiency defect $\lambda$, presented in the work of Kotas \cite{Kotas2013}, defined as the fraction of the total exergy input destroyed in the successive irreversible processes
	\begin{equation} \lambda = \frac{\dot{E}_{\mathrm{d}}}{\dot{E}_{\mathrm{in,tot}}} \end{equation}
	\item  the irreversibility share $\delta$, suggested in the work of Tsatsaronis \cite{Tsatsaronis1993}, defined as the ratio between the exergy destroyed in the $i$-th component in relation to the exergy destroyed in the entire system
	\begin{equation} \delta_i = \frac{\dot{E}_{\mathrm{d},i}}{\dot{E}_{\mathrm{d,tot}}} \end{equation}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.95\linewidth]{Figures/SystemGeneral}
	\caption{Layout of the case study energy systems}
	\label{fig:shipSystems}
\end{figure}

@@

The operational data was collected on board from the ships' machine logging and surveillance system. The data sampling rate was selected to 15 minutes, automatically calculated as averages over the respective periods by the on board data logging systems, as an acceptable compromise between accuracy and computational time required for the data processing. Data consistency was checked individually by creating a descriptive statistic and a histogram of each measurement. A filter was applied by setting up a pre-defined maximum and minimum value.

Since measurements of ambient and seawater temperature from onboard logging systems were not available for the whole dataset, we used measurements taken from SMHI database for Landsort lighthouse for the seawater, and the lighthouse Svenska H\"{o}garna for the ambient temperature, both located along the route of the vessel.

@@

The main principles applied to the modeling are presented in the following sections and summarized in Figure \ref{fig:method}, while a more thorough description and validation of the different approaches is presented in the Appendix.

\begin{figure}
	\centering
	\includegraphics[width=0.95\linewidth]{Figures/dataProcessing}
	\caption{Overview of the data processing from inputs to outputs}
	\label{fig:dataprocessing}
\end{figure}

@@

The Diesel engines are the only power source of the system, subdivided in four main engines uniquely devoted to propulsion and four auxiliary engines used for on board electric power generation. The energy and exergy analysis of the engines is not limited to the boundary of the system, but is brought to the detail of the different components that constitute the engine systems, as shown in Figure \ref{fig:DieselEngines}.

In the case of the main engines, the engine power output was not measured and needed to be estimated. In this work, we calculated the engine power output based on measurements of engine fuel rack position (used as a proxy of the mass of fuel injected per cycle) and of the engine speed (see equations \ref{eqn:ME:eq1} to \ref{eqn:ME:eq3}). In the case of the auxiliary engines, instead, direct measurements of the power generated by each engine were available. 
\begin{eqnarray}
\dot{m}_{fuel} & = & \dot{m}_{fuel,des} \left(a_0 + a_1 \frac{frp}{frp_{des}}\right) \frac{\omega}{\omega_{des}} \label{eqn:ME:eq1} \\
\eta_{ME} & = & a_0 + a_1 \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} + a_2 \left( \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} \right)^2 \label{eqn:ME:eq2} \\
\dot{W}_{ME} & = & \dot{m}_{fuel} \eta_{ME} LHV \label{eqn:ME:eq3}
\end{eqnarray}

@@

For both auxiliary and main engines, part of the relevant internal variables are not measured (particularly with relation to the bypass valves) and this required to determine them based on modeling the heat and mass balance of the engine. Values for temperatures, mass flows and energy flows for different engine loads are provided in the \ref{sec:appendix:engines}, for both main and auxiliary engines.

@@

The total electric power demand of the system is easily estimated as the sum of the power generated by the four auxiliary engines, that is directly measured on the electrical side of the generators and has a rather high accuracy.
\begin{equation}
P_{el,tot} = \sum_{1}^{4} P_{AE_i}
\end{equation}

The ship is equipped with a number of different systems, including lighting, navigational systems, pumps and compressors, etc. However, none of the individual contributions is directly measured, hence estimations of the energy demand of individual consumers need to be based on indirect measurements. In this paper only the contributions of the HVAC and the bow thrusters are identified separately from the remaining parts of the system. More details about the methods employed in this process are provided in Appendix REF.

\subsection{Heating and cooling systems} \label{sec:met:heat}

@@

The ship's cooling systems are designed similarly to most ships and divided in two cooling systems operated at two separate temperature levels: the high temperature (HT) cooling systems (70-90$^o$C) and the low temperature (LT) cooling systems (40-60$^o$C). The cooling systems provide the necessary cooling to the engines, to the steam systems and to all other components on board. An overview of the ship heating and cooling systems is provided in Figure \ref{fig:generalsystemssimplified}.

All values not directly measured in the cooling systems were calculated based on mass and energy balances. Mass flows were determined based on measurements of the pressure in the system and on the performance curves of the cooling pumps as provided in the engine project guides.
The main view of the ship heating and cooling systems is presented in Figure \ref{fig:ship:general}.

@@

ADD TABLE WITH LEGEND

The determination of the heat demand is the most challenging in the analysis of ship energy systems, as extensive and appropriate measurements related to this contribution are rarely available. In this case the limited data available was fit with assumptions related to the behavior of the energy system and to the daily repartition of the heating demand. A more detailed description of this procedure is provided in Appendix REF.

@@

Improving the performance of a ship requires a detailed analysis of the energy demands to optimize the dimensioning and operation of the technologies and processes to install on-board. It requires the analysis of the hourly, daily and seasonal variations of the heat/power consumption, operating (e.g. ship speed) and external conditions (e.g. air temperature). However, gathering and processing those data may be time-consuming because of the high number of points to assess. It is therefore desirable to reduce this large quantity of information while keeping the relevant details on the relations between the variables of interest. 
One way is to represent these yearly profiles in a limited number of representative (typical) periods, avoiding the repetition of similar data sets. Clustering consists of grouping such sets in a single group (cluster) so that the items in the same group are more like each other than to those belonging to other groups. Several algorithms are introduced in the literature to partition these datasets. The one selected in this work is the Lloyd’s (k-means) algorithm \cite{Kanungo2002}, which is a non-deterministic method computationally faster than conventional hierarchical tools. The approach applied by Fazlollahi et al. \cite{Fazlollah2014} was used to select the optimum number of clusters (Nk). It (Nk) should be as low as possible for data handling purposes, while preserving a high accuracy of the retrieved data. It builds on the calculation of three criteria: 
\begin{enumerate}
	\item the average intra-cluster distance to assess the density of each cluster, preferably small;
	\item the average inter-cluster distance to assess the distance between each, preferably high; 
	\item the expected square error, which is a statistical measure suggested by Pham et al.. 
\end{enumerate}

In parallel, the cluster quality was systematically evaluated by calculating the following performance indicators:
\begin{enumerate}
	\item the profile deviation (deviation between original and typical period profiles), $\sigma_{profile}$;
	\item the deviation from the load duration curve of the average values of each period, $\sigma_{CDC}$;
	\item the relative error in load duration curve deviation $ELDC$;
	\item the maximum duration load curve difference $\Delta_{MLDC}$;
	\item the number of periods with relative errors $\Delta_{prod}$ higher than 7\%.
\end{enumerate}

@@

Figure \ref{fig:vship_hist} represents the distribution of the ship speed. The ship operates for almost  the entire year according to a fixed schedule, while there are a couple of periods (particularly during the summer months) when the ship operates at higher speed. In Figure \ref{fig:opMode_pie} it can be also observed that the ship spends a relatively large amount of time in port. These remarks suggest that, in spite of the high installed power of the main engines, the energy demand for propulsion is expected to be relatively low.

@@

The ship energy demand is first subdivided among the different types of consumers. The evolution of the demands of propulsion, electric power, and heat over time during a typical voyage is shown in Figures \ref{fig:Demand_vs_time_W} and \ref{fig:Demand_vs_time_S}, representing a winter and a summer day respectively. As expected, the heating demand in winter is higher than in summer [2800-4500 kW vs 1100-2900 kW], as a consequence of the reduced need for compartment heating. On the other hand, the electric power demand behaves inversely, ranging around 1900 kW during the reference winter day (peaks are connected to the use of thrusters in port for maneuvering) and around 2300 kW during the reference summer day, where the difference is mostly associated to the demand of the HVAC compressors. Looking at the yearly cumulated demand, nearly 50\% of the total energy demand is related to ship propulsion (see Figure \ref{fig:Pie_EnergyDemand}), while the remaining portion is approximately equally split between electric energy and heat demand.

@@

Heat recovery has a large impact on the overall external heat demand (see Fig.  \ref{fig:Pie_HeatGeneration}). The exhaust gas boilers and the heat recovery on the HT-water fulfill almost 75\% of the yearly demand for heating, leaving only the remaining 25\% to be provided by oil-fired auxiliary boilers. The contribution of the HRSG and the HTHR is substantially constant throughout the year (Fig. \ref{fig:TimeSeries_HeatGeneration}), while the auxiliary boilers are used to cover the peaks and the winter increased baseload.

@@

Given the fact that many engines are operated at low load (both main and auxiliary engines), the system would benefit from both an electrification of the system and from the installation of batteries. The system electrification was explored by the authors \cite{Baldi2016b} and showed a relevant potential from both an environmental and an economic point of view. The installation of batteries was also considered in a previous publication by the authors \cite{Baldi2017} and proved the potential for yearly savings of 1-2\%. In this second case, however, it should be noted that the economic performance was not evaluated quantitatively.

The availability of waste heat that is not already recovered by the system suggests that the system could benefit from the installation of a Heat-to-power WHR. This possibility was tested by the authors both in the case that no additional retrofit is performed \cite{Mondejar2017,Ahlgren2016}, with estimated savings of 22\% of the auxiliary power demand for a reference voyage, and in the case of a full integration with the rest of the system \cite{Baldi2017}, with estimated savings of up to 6\% of the total fuel consumption for a year of ship operations.

@@

\section{Energy system modelling}

\subsection{Engine modelling, figures and tables} \label{sec:appendix:engines}

\begin{align}
\dot{m}_{air,comp} & =  \dot{m}_{air,cyl} + \dot{m}_{air,bp} \label{eq:mb:comp} \\ 
\dot{m}_{eg,turb} & = \dot{m}_{eg,cyl} + \dot{m}_{air,bp} \label{eq:mb:turb} \\
\dot{m}_{eg,cyl} & = \dot{m}_{air,cyl} + \dot{m}_{fuel,cyl} \label{eq:mb:cyl} \\
\dot{m}_{air,comp} \Delta h_{comp} & = \dot{m}_{eg,turb} c_{p,eg} (T_{turb,in} - T_{turb,out}) \eta_{mech} \label{eq:eb:turbocharger} \\
\dot{m}_{eg,turb} c_{p,eg} (T_{turb,in} - T_0) & = \dot{m}_{air,bp} c_{p,air} (T_{comp,out} - T_0) + \dot{m}_{eg,cyl} c_{p,eg}  (T_{cyl,out} - T_0) \label{eq:eb:bpmerge}
\end{align}
Where equations \ref{eq:mb:comp} to \ref{eq:mb:cyl} represent the mass balances of the bypass split, bypass merge, and cylinder respectively, while equations \ref{eq:eb:turbocharger} and \ref{eq:eb:bpmerge} represent the energy balances of the turbocharger and of the bypass merge, respectively.

@@

The most relevant operative values are provided in Tables \ref{tab:ME_values} and \ref{tab:AE_values} and the main energy flows are listed in Tables \ref{tab:ME_flows} and \ref{tab:AE_flows}.

@@

\subsection{Electric demand modelling}
